An efficient acyclic contact planner for multiped robots

### 2017-01-11 body/contact.tex
[Contact planning] + cosmetics in overview intro and rbprm

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Equil}^0$}.
If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.

In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
root location.
Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.

Our criterion to assert efficiently the static equilibrium of the system
is described in Section~\ref{sec:heuristics}.

\subsection{Definition of a contact generator}
\label{sec:single_contact}
Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
$R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
\begin{enumerate}
\item Generate randomly a collision-free limb configuration;
\item Project the end-effector onto the closest surface with inverse kinematics;
\item If a valid solution is found, stop. Otherwise repeat from step 1.
\end{enumerate}

We trade the completeness of this approach for a more efficient solution introduced in our previous work~\citep{Tonneau2014}, which we describe
here to be exhaustive.

We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimum distance 
between an effector and a surface of the environment is less than a small value $\epsilon$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item Select the first configuration of $S$. Project the configuration onto contact with inverse kinematics. If $S$ is empty, stop (failure case) (Figure~\ref{fig:contact_gen} (d) and (e));
\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back to step 5.
\end{enumerate}

This approach has two main advantages.
First, this allows us to select a large number of candidates in a single proximity request.
Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
a locally-optimal candidate.
Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
$N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.

@@

\subsection{Extension of the guide path}
Using the contact generator, we define an iterative algorithm to generate the contact
sequence along the guide path.
Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
the contacts are created, allowing it to break the combinatorial.
The complete algorithm can be found in Appendix~\ref{app:contact}.
In the remainder of the section we provide an intuition of it.

As an input, we consider the guide path $\mathbf{q}^0(t)$, discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations. %To ensure continuity in the contact transition phases,
We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
in \gls{ccequil} such that contacts are maintained if possible.
The first full body configuration of the sequence is given by the initial state of the robot.

@@

Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
If possible, a limb in contact at step $i-1$ remains in contact at step $i$. The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. The solver is directly provided by the HPP software.

If the solver fails, the contact is broken and a collision-free configuration is assigned to the limb.
If two or more contacts are broken in a single step, one or more intermediate steps are added.
For these steps the root configuration is the same as for the previous step, with the difference that
one faulty contact is repositioned, in the hope that it will not be broken at the next step.

@@

%\subsection{Generation of a contact configuration}  
\subsubsection{Creating contacts:}
Contacts are created using three rules:
\begin{enumerate}
\item Only one contact creation can happen between two consecutive steps. This increases the odds that the interpolation between the two steps be feasible;
%\adnote{Before you said that you ensured that transition between two states is feasible...}
\item A contact is validated if and only if the resulting configuration is in static equilibrium; 
\item We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.
\end{enumerate}

These three rules provide a deterministic generation of contacts along the discretized guide path.
For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.

>>>>

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
In the second phase, we compute a discrete sequence of static equilibrium configurations $\mathbf{Q}^{\overline{0}}$ given a root path
$\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Reach}^0 \simeq C_{Equil}^0$}. This contact planner is built on top of
a contact generator, used to generate static equilibrium configurations. We first describe the generator, before detailing the contact planning algorithm.
%~ Our planner computes guide paths in $C_{Reach}^0$ , an approximation of \gls{$C_{Equil}^0$}.
%~ As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{$C_{Equil}^0$}.
%~ If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
%~ We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.

%~ In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
%~ root location.
%~ Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.

%~ Our criterion to assert efficiently the static equilibrium of the system
%~ is described in Appendix~\ref{sec:heuristics}.

\subsection{Definition of a contact generator}
\label{sec:single_contact}

\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
		\put (1,1) {a} 
		\put (22,1) {b} 
		\put (42,1) {c} 
		\put (62,1) {d} 
		\put (83,1) {e} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right leg of HRP-2. (a): Selection of reachable obstacles. (b): Entries of the limb samples database (with $N = 4$). (c): With a proximity query on the octree database, configurations too far from obstacles are eliminated. (d): The best candidate according to a user-defined heuristic $h$ is chosen. (e): The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure*}

A contact generator is defined for a given limb $R^k$, and only affects the dofs of $R^k$.
Given a robot configuration, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
\begin{equation}
\label{eq:contact_gen}
	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \wedge \mathbf{q}^k \in  C_{Contact}^k 
\end{equation}

Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the considered effector onto the closest surface with an inverse kinematics solver.
This process is repeated until either a solution for Eq.~\ref{eq:contact_gen} is found or the generator fails, according to a user-defined maximum number of trials.
We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.

%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
%~ \begin{enumerate}
%~ \item Generate randomly a collision-free limb configuration;
%~ \item Project the end-effector onto the closest surface with inverse kinematics;
%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
%~ \end{enumerate}

%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.

We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
between an effector and an obstacle is lesser than $\epsilon \in \mathbb{R}$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
\item Use a user-defined heuristic $h$ to sort $S$;
\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics. (Figure~\ref{fig:contact_gen} (d) and (e));
\item If Eq.~\ref{eq:contact_gen} is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
\end{enumerate}

%~ This approach has two main advantages.
%~ First, this allows us to select a large number of candidates in a single proximity request.
%~ Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
%~ a locally-optimal candidate.
%~ Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
%~ Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
%~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
of the $N$ parameter \citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ used are detailed in Appendix~\ref{sec:heuristics}.

@@

\subsection{Contact planning algorithm}
Starting from an initial whole-body configuration, we use the contact generator to compute a sequence
of whole-body configurations  $\mathbf{Q}^{\overline{0}}$ along the root path $\mathbf{q}^0(t)$.
The algorithm details can be found in Appendix~\ref{app:contact}. Here we provide an intuition of it.
%~ Although we address acyclic contact sequences, the algorithm is deterministic in the order in which 
%~ the contacts are created, allowing it to break the combinatorial.

First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:
\begin{itemize} 
\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
\item Each configuration is in static equilibrium.
\item Each configuration is collision-free.
\end{itemize}

%~ We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
%~ To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
%~ in $C_{Equil}$ such that contacts are maintained if possible.
%~ The first full body configuration of the sequence is given by the initial state of the robot.

@@

%~ Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
If kinematically possible, a limb in contact at step $i-1$ remains in contact at step $i$ (Figure~\ref{fig:break_contact}). 
%~ The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. 
%~ The solver is directly provided by the HPP software.
Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
If two or more contacts are broken in a single step, one or more intermediate steps are added.
%~ For these steps the root configuration is the same as for the previous step, with the difference that
%~ one faulty contact is repositioned, in the hope that it will not be broken at the next step.

@@

%\subsection{Generation of a contact configuration}  
\subsubsection{Creating contacts:}
Contacts are created using three rules:
\begin{enumerate}
\item At most one contact creation happens between two consecutive steps. 
%~ This increases the odds that the interpolation between the two steps be feasible;
%\adnote{Before you said that you ensured that transition between two states is feasible...}
\item A contact is validated if and only if the resulting configuration is in static equilibrium; 
\item We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.
\end{enumerate}

The set of rules defined in this Section define a deterministic contact planning algorithm.
The determinism of this algorithm makes it efficient, at the cost of not being probabilistically complete.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.

### 2017-01-11 body/overview.tex
[Contact planning] + cosmetics in overview intro and rbprm

%
\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
		\put (1,1) {a} 
		\put (22,1) {b} 
		\put (42,1) {c} 
		\put (62,1) {d} 
		\put (83,1) {e} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right leg of HRP-2. (a): Selection of reachable obstacles. (b): Entries of the limb samples database (with $N = 4$). (c): With a proximity query on the octree database, configurations too far from obstacles are eliminated. (d): The best candidate according to a user-defined heuristic $h$ is chosen. (e): The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure*}

@@

An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a) to allow contact creation; not too close, because the robot must avoid collision.
%~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision, while the reachable workspace is in collision.
%~ Verifying the \textit{reachability condition} is very efficient in practice.
To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
This is detailed in Section~\ref{rbprm}.
%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).

@@

The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
To achieve this the root path is first decomposed into a sequence of discrete root configurations, according to a user-defined discretization step.
Each root configuration is then extended into a whole-body configuration, subject to three constraints:
\begin{itemize} 
\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
\item Each configuration is in static equilibrium.
\item Each configuration is collision-free.
\end{itemize}

The algorithm thus proceeds iteratively, starting from the whole-body initial configuration of the robot.

The algorithm takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually. Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
  \caption{
           Reachable workspace and torso bounding box of HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$.}
		   \label{fig:HyQ_roms}
\end{figure}

@@

\medskip
$C_{Contact}$ is the set of whole body-configurations in contact and collision-free.

@@

%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.

>>>>

%

@@

An intuitive description of \gls{contact reachable} configurations is ``close, but not too close'' to the environment: close, because a contact surface must be partially included in the reachable workspace of the robot to allow contact creation; not too close, because the robot must avoid collision.
%~ \deladp{\textit{Contact reachability} is verified with the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:reach_cond}).}
More precisely, a root configuration is \textit{contact reachable} if the root scaled by a user-defined factor $s \geq 1$ is not in collision (Figure~\ref{fig:reach_cond} - red shape), while the reachable workspace is in collision (Figure~\ref{fig:reach_cond} - green shapes).
%~ Verifying the \textit{reachability condition} is very efficient in practice.
To plan a root path, we then use an RRT planner where, instead of checking collision to validate a configuration, we verify the \textit{reachability condition}.
This is detailed in Section~\ref{rbprm}.
%~ In Section~\ref{rbprm} we detail how  we define and use the \textit{reachability condition} to compute a guide path with the Reachability-Based RRT (Figure~\ref{fig:framework}---\Pa).

@@

The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--$\mathcal{P}_2$).
To achieve this the root path is first decomposed into a sequence of discrete root configurations, according to a user-defined discretization step.
Each root configuration is then extended into a whole-body configuration in static equilibrium.
The algorithm thus proceeds iteratively, starting from the whole-body initial configuration of the robot.
It takes advantage of the fact that each root configuration is fixed to generate the contact by considering each limb individually. 
%~ Candidate limb configurations resulting in the end-effector being close to the environment are selected from a database computed offline independently from the environment).
%~ Then we select one of these configurations based on user-defined heuristics (Figure~\ref{fig:contact_gen}), detailed in Appendix~\ref{sec:heuristics}.

@@

@@

\medskip
$C_{Contact}$ is the set of whole body-configurations in contact and collision-free.
$C_{Contact}^k \subset C_{Contact}$ is the set of whole body-configurations where at least $R^k$ is in contact.

@@

%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.
\begin{figure}
  \centering
  \includegraphics[width=0.95\linewidth]{figures/HyQ_roms}
  \caption{
           Reachable workspace and torso bounding box of HyQ. The green shapes represent the reachable workspace $W^k$ of each limb. The red shape is $W^0$.}
		   \label{fig:HyQ_roms}
\end{figure}
